Pushdown Control-Flow Analysis via Refunctionalization (Functional Pearl)

### 2018-02-25 paper.tex
update

%% Abstract
%% Note: \begin{abstract}...\end{abstract} environment must come
%% before \maketitle command
\begin{abstract}
  Abstracting abstract machines (AAM) is a systematic methodology 
  for constructing abstract interpreter that derived from a concrete
  small-step abstract machine. Recent progress applies the same idea 
  on concrete definitional interpreter and obtains a big-step 
  abstracting definitional interpreter (ADI) written in monadic style.
  Danvy shows the correspondence between abstract machines and interpreters.
  In this paper, we study the relations between small-step AAM and 
  big-step ADI.

@@

Tranditionally, continuation-passing style (CPS) is a popular intermediate representation
for analyzing functional programs because of it exposes control transfer explicitly
and simplifies the analysis \cite{Shivers:1991:SSC:115865.115884, Shivers:1988:CFA:53990.54007}.
Here, we choose to use administrative normal form (ANF) which is a direct-style intermediate representation 
as target for clarity but without losing simplicty and generality.
The transformations we will show in the rest of this paper
also work on abstract machines for plain, direct-style $\lambda$-calculus languages.

@@

\begin{verbatim}
e \in Exp ::= ae
            | (let ([x (f ae)]) e)
            | (letrec ([x ae] ...) e) (TODO: maybe not include letrec?)
ae \in AExp ::= x
              | lam
lam \in Lam ::= (lambda (x) e)
x ... are variable names
\end{verbatim}

@@

case class Var(x: String) extends Expr
case class App(e1: Expr, e2: Expr) extends Expr
case class Lam(x: String, body: Expr) extends Expr
case class Let(x: String, e: App, body: Expr) extends Expr
case class Letrec(bds: List[Binding], body: Expr) extends Expr
\end{verbatim}

@@

To run the program, we use function \texttt{inject} to construct an initial machine
state given an expression \texttt{e}.
Then the \texttt{drive} function is used to evaluate
to a final state by iteratively apply \texttt{step} on current state until we reach a state
that the control is an atomic expression and the continuation stack is empty.
And of course, we can extract the value from the final state at last.

\begin{verbatim}
def drive(s: State): State = s match {
  case State(ae, _, _, Nil) if isAtomic(ae) => s
  case _ => drive(step(s))
}

def inject(e: Expr): State = State(e, Map(), Map(), Nil)
def eval(e: Expr): State = drive(inject(e))
\end{verbatim}

@@

The co-domain of binding stores is \texttt{Storable} which is the same
as we defined for CESK machines.
The co-domain of continuation stores \texttt{Cont} is comprised of 
a frame \texttt{Frame} and a continuation address \texttt{KAddr}.
To mimic the runtime call stack, \texttt{KAddr} plays the role of
representing the remaining stack frames.
TODO: nondet of cont?

@@

Up to now, we did not describe much about the time-stamp \texttt{Time}
os states and addresses of stores.
But they are key ingredients to achieve analysis with different sensitivities,
as well as a finite state space analysis.
To effectively approximate the runtime behavior, a finite 
program contour \texttt{time} that encodes the program execution history is introduced.
Here, we use a list of execution contexts (expressions) to reprent that, and
as we will see in section \ref{kcfainst}, by using different \texttt{tick}
function on time we may obtain a family of analyses.

@@

Address.

@@

@@

To allocate addresses for bindings or continuations, we use \texttt{allocBind}
and \texttt{allocKont}.

\begin{verbatim}
def allocBind(x: String, time: Time): BAddr = BAddr(x, time)
def allocKont(tgtExpr: Expr, tgtEnv: Env, 
              tgtStore: BStore, time: Time): KAddr = ContAddr(tgtExpr, time)

@@

\begin{verbatim}
def step(s: State): List[State] = {
  val new_time = tick(s)
  s match {
    case State(Let(x, App(f, ae), e), env, bstore, kstore, kaddr, time) 
      if isAtomic(ae) =>
      val closures = atomicEval(f, env, bstore).toList
      for (Clos(Lam(v, body), env_c) <- closures) yield {
        val baddr = allocBind(v, new_time)
        val new_env = env_c + (v -> baddr)
        val new_bstore = bstore.update(baddr, atomicEval(ae, env, bstore))
        val new_kaddr = allocKont(body, env_c, new_bstore, new_time)
        val new_kstore = kstore.update(new_kaddr, Cont(Frame(x, e, env), kaddr))
        State(body, new_env, new_bstore, new_kstore, new_kaddr, new_time)
      }
    case State(ae, env, bstore, kstore, kaddr, time) 
      if isAtomic(ae) =>
      val conts = kstore(kaddr).toList
      for (Cont(Frame(x, e, env_f), f_kaddr) <- conts) yield {
        val baddr = allocBind(x, new_time)
        val new_env = env_f + (x -> baddr)
        val new_store = bstore.update(baddr, atomicEval(ae, env, bstore))
        State(e, new_env, new_store, kstore, f_kaddr, new_time)
      }
  }
}
\end{verbatim}

@@

Comparing with concrete CESK machine, the
\texttt{drive} function performs a collecting semantics rather than valuation
semantics. That is, for the purpose of analyzing a program, \texttt{drive}
collects all the intermediate machine states as the program is abstractly executing.
The following code shows a variant of work list algorithm that always applies \texttt{step}
function to the first element of \texttt{todo} and adds it to \texttt{seen} set
if there is an unexplored one. If the work list is empty, the function just
returns the set of explored states up to now.

\begin{verbatim}
def inject(e: Expr): State = 
  State(e, Map(), 
        Store[BAddr, Storable](Map()), 
        Store[KAddr, Cont](Map(Halt -> Set())), 
        Halt, List())

def drive(todo: List[State], seen: Set[State]): Set[State] = todo match {
  case Nil => seen
  case hd::tl if seen.contains(hd) => drive(tl, seen)
  case hd::tl => drive(step(hd).toList ++ tl, seen + hd)
}

@@

@@

In this section, we firstly describe a naive and uncomputable solution to the
problem; then briefly review several recent research works that enable
computable and precise call/return match. Even though the first solution is
uncomputable, but as we will see in section 3, it is the starting point of our
transformation.

@@

Recent years, there are great efforts \cite{vardoulakis2010cfa2, earl2012introspective, gilray2016pushdown}
to achieve precise call/return match based
on small-step abstracting abstract machines which is what we described in section \ref{aam}.

go over existing computable solutions P4F ANF, not works on direct style,PDCFA,AAC

@@

\section{Detour: Shiver's \textit{k}-CFA}

>>>>

%% Abstract
%% Note: \begin{abstract}...\end{abstract} environment must come
%% before \maketitle command
\begin{abstract}
  Abstracting abstract machines (AAM) is a systematic methodology 
  for constructing abstract interpreters that derived from concrete
  small-step abstract machines. Recent progress applies the same idea 
  on definitional interpreter and obtains a big-step 
  abstracting definitional interpreter (ADI) written in monadic style.
  Danvy shows the correspondence between abstract machines and interpreters.
  In this paper, we study the relations between small-step AAM and 
  big-step ADI.

@@

Tranditionally, continuation-passing style (CPS) is a popular intermediate representation
for analyzing functional programs because of it exposes control transfer explicitly
and simplifies the analysis \cite{Shivers:1991:SSC:115865.115884, Shivers:1988:CFA:53990.54007}.
Here, we choose to use administrative normal form (ANF) which is a direct-style intermediate representation 
as target for clarity but without losing simplicty and generality.
Although we just show the core calculus language, but it can be easily extended
to support recursive bindings, conditionals, primitive types and operations on primitive types.
The transformations we will show in the rest of this paper
also work on abstract machines for plain, direct-style $\lambda$-calculus languages.
TODO: add letrec?

@@

\begin{verbatim}
e \in Exp ::= ae
            | (let ([x (f ae)]) e)
ae \in AExp ::= x
              | lam
lam \in Lam ::= (lambda (x) e)
x ... are variable names
\end{verbatim}

@@

case class Var(x: String) extends Expr
case class App(e1: Expr, e2: Expr) extends Expr
case class Lam(x: String, body: Expr) extends Expr
case class Let(x: String, e: App, body: Expr) extends Expr
\end{verbatim}

@@

To run the program, we first use function \texttt{inject} to construct an initial machine
state given an expression \texttt{e}. The initial state contains an emtpy environment, 
store and stack.
\begin{verbatim}
def inject(e: Expr): State = State(e, Map(), Map(), Nil)
\end{verbatim}

Then the \texttt{drive} function is used to evaluate
to a final state by iteratively apply \texttt{step} on current state until we reach a state
that the control is an atomic expression and the continuation stack is empty.
And of course, we can extract the value from the final state at last.

\begin{verbatim}
def drive(s: State): State = s match {
  case State(ae, _, _, Nil) if isAtomic(ae) => s
  case _ => drive(step(s))
}
def eval(e: Expr): State = drive(inject(e))
\end{verbatim}

@@

The co-domain of binding stores \texttt{Storable} is the same
as we defined for CESK machines.
The co-domain of continuation stores \texttt{Cont} is comprised of 
a frame \texttt{Frame} and a continuation address \texttt{KAddr}.
To mimic the runtime call stack, \texttt{KAddr} plays the role of
representing the remaining stack frames.
TODO: nondet of cont?

@@

\subsubsection{Allocating Addresses}
Up to now, we did not describe much about the adresses in stores and the 
time-stamp \texttt{Time}.
But they are key ingredients to achieve analyses with different sensitivities,
as well as a finite state space analysis.
To effectively approximate the runtime behavior, a finite 
program contour \texttt{time} that encodes the program execution history is introduced.
Here, we use a list of execution contexts (expressions) to reprent that, and
as we will see in section \ref{kcfainst}, by using different \texttt{tick}
function on time we may obtain a family of analyses.

@@

As we mentioned previously, the space of addresses is finite in AAM.
Binding addresses are parameterized by variable names and the time that creates this binding.
Continuation addresses \texttt{KAddr} has two variants: 
1) \texttt{Halt} which corresponds to the empty stack, and
2) \texttt{ContAddr} consists of the target expression of callee and the time that
creates this address. (TODO: why target expression, allocation polyvariance)

@@

We use two helper functions \texttt{allocBind} and \texttt{allocKont} that will be used later
to allocate binding addresses and continuation addresses.

\begin{verbatim}
def allocBind(x: String, time: Time): BAddr = BAddr(x, time)
def allocKont(tgtExpr: Expr, time: Time): KAddr = ContAddr(tgtExpr, time)
\end{verbatim}

@@

atomicEval returns a set of storable values, i.e., closures.

\begin{verbatim}

@@

\begin{verbatim}
def step(s: State): List[State] = {
  val new_time = tick(s)
  s match {
    case State(Let(x, App(f, ae), e), env, bstore, kstore, kaddr, time) 
      if isAtomic(ae) =>
      val closures = atomicEval(f, env, bstore).toList
      for (Clos(Lam(v, body), env_c) <- closures) yield {
        val baddr = allocBind(v, new_time)
        val new_env = env_c + (v -> baddr)
        val new_bstore = bstore.update(baddr, atomicEval(ae, env, bstore))
        val new_kaddr = allocKont(body, new_time)
        val new_kstore = kstore.update(new_kaddr, Cont(Frame(x, e, env), kaddr))
        State(body, new_env, new_bstore, new_kstore, new_kaddr, new_time)
      }
    case State(ae, env, bstore, kstore, kaddr, time) 
      if isAtomic(ae) =>
      val conts = kstore(kaddr).toList
      for (Cont(Frame(x, e, env_f), f_kaddr) <- conts) yield {
        val baddr = allocBind(x, new_time)
        val new_env = env_f + (x -> baddr)
        val new_store = bstore.update(baddr, atomicEval(ae, env, bstore))
        State(e, new_env, new_store, kstore, f_kaddr, new_time)
      }
  }
}
\end{verbatim}

@@

To run (analyze) a program, as CESK machine, we use \texttt{inject} to construct
the initial state given a program. 
Note that the inital continuation store has a special mapping that maps continuation address
\texttt{Halt} to empty set of continuations.
We also provide an empty program contour as inital time.
\begin{verbatim}
def inject(e: Expr): State = 
  State(e, Map(), 
        Store[BAddr, Storable](Map()), 
        Store[KAddr, Cont](Map(Halt -> Set())), 
        Halt, List())
\end{verbatim}

However, comparing with concrete CESK machine, the
\texttt{drive} function performs collecting semantics instead of valuation
semantics. That is, for the purpose of analyzing programs, the function \texttt{drive}
collects all the intermediate machine states as the program is abstractly executing.
The following code shows a variant of work list algorithm that always applies \texttt{step}
function to the first element \texttt{hd} of the work list \texttt{todo} if \texttt{hd}
is unseen, then adds the result of \texttt{step} to the work list and also adds \texttt{hd}
to the explored states set.
If the work list is empty, the function just
returns the set of explored states up to now.

\begin{verbatim}
def drive(todo: List[State], seen: Set[State]): Set[State] = todo match {
  case Nil => seen
  case hd::tl if seen.contains(hd) => drive(tl, seen)
  case hd::tl => drive(step(hd).toList ++ tl, seen + hd)
}

@@

Finally, a user could use function \texttt{analyze} to obtain all possible states 
for a given program.

@@

In this section, we firstly describe a simple but uncomputable solution to the
problem; then briefly review several recent research works that enable
computable and precise call/return match. Even though the first solution is
uncomputable, but as we will see in section 3, it is the starting point of our
transformation.

@@

Recent years, there are great efforts \cite{vardoulakis2010cfa2, earl2012introspective, 
gilray2016pushdown, johnson2015abstracting}
to achieve precise call/return match based on small-step AAM which is what we 
described in section \ref{aam}.

TODO: summarize CFA2, PDCFA, AAC, P4F

On the other side, \citeauthor{darais2017abstracting} construct abstracting 
definitional interpreters (ADI) that automatically inherit the pushdown control-flow 
property from the defining language. TODO

In the rest of this section, we take a closer look at the two state-of-art techniques
to accomplish perfect call-stack precision.

\textbf{Pushdown for Free.}
\citeauthor{gilray2016pushdown} proposed a continuation address allocator for 
small-step abstracting abstract machine that is both easy to implement and
computational inexpensive.

\begin{verbatim}
case class P4FContAddr(tgt: Expr, tgtEnv: Env, time: Time) extends KAddr
def allocKont(tgtExpr: Expr, tgtEnv: Env, time: Time): KAddr = 
  P4FContAddr(tgtExpr, tgtEnv, time)
\end{verbatim}

\textbf{Abstracting Definitional Interpreters.}

@@

\section{Detour: Shiver's \textit{k}-CFA?}
